---
title: "Health status by age"
vignette: >
  %\VignetteIndexEntry{Health status by age}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 0.35*7
)
```


For this example, we will use the following packages.
```{r message = FALSE}
library(tidyverse)
library(cowplot)   # for theme_minimal_hgrid()
```

The dataset is provided as `practicalgg::happy`. Let's look at it in table form and as basic density plots.

```{r message = FALSE}
data_health <- practicalgg::happy %>%
  select(age, health) %>%
  na.omit() %>%
  mutate(health = fct_rev(health)) # revert factor order

data_health

ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(fill = "lightblue") +
  facet_wrap(~health, nrow = 1)
```

Add the overall distribution as a background.
```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  # we add a density layer that uses the data without the column
  # by which we are faceting; this places the full dataset into
  # each facet
  geom_density(
    data = select(data_health, -health),
    aes(fill = "all people surveyed"),
  ) +
  geom_density(aes(fill = "highlighted group")) +
  facet_wrap(~health, nrow = 1)
```

Define the scales.
```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(
    data = select(data_health, -health),
    aes(fill = "all people surveyed"),
  ) +
  geom_density(aes(fill = "highlighted group")) +
  scale_x_continuous(
    name = "age (years)",
    limits = c(15, 98),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "count",
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#2b8cbed0"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  facet_wrap(~health, nrow = 1)
```

Basic theme; move legend to bottom; remove outline around densities.
```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(
    data = select(data_health, -health),
    aes(fill = "all people surveyed"),
    color = NA
  ) +
  geom_density(aes(fill = "highlighted group"), color = NA) +
  scale_x_continuous(
    name = "age (years)",
    limits = c(15, 98),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "count",
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#2b8cbed0"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  facet_wrap(~health, nrow = 1) +
  theme_minimal_hgrid(12) +
  theme(
    legend.position = "bottom",
    legend.justification = "right"
  )
```

Theme tweaks: Larger strip labels, move legend closer to plot, adjust horizontal legend spacing.

```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(
    data = select(data_health, -health),
    aes(fill = "all people surveyed"),
    color = NA
  ) +
  geom_density(aes(fill = "highlighted group"), color = NA) +
  scale_x_continuous(
    name = "age (years)",
    limits = c(15, 98),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "count",
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#2b8cbed0"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  facet_wrap(~health, nrow = 1) +
  theme_minimal_hgrid(12) +
  theme(
    strip.text = element_text(size = 12, margin = margin(0, 0, 6, 0, "pt")),
    legend.position = "bottom",
    legend.justification = "right",
    legend.margin = margin(6, 0, 1.5, 0, "pt"),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(0, "pt"),
    legend.box.spacing = grid::unit(0, "pt")
  )
```

Remove axis line, add spacing between legend items.

```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(
    data = select(data_health, -health),
    # a simple workaround to a limitation in ggplot2:
    # add a few spaces at the end of the legend text
    # to space out the legend items
    aes(fill = "all people surveyed   "),
    color = NA
  ) +
  geom_density(aes(fill = "highlighted group"), color = NA) +
  scale_x_continuous(
    name = "age (years)",
    limits = c(15, 98),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "count",
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#2b8cbed0"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  facet_wrap(~health, nrow = 1) +
  theme_minimal_hgrid(12) +
  theme(
    axis.line = element_blank(),
    strip.text = element_text(size = 12, margin = margin(0, 0, 6, 0, "pt")),
    legend.position = "bottom",
    legend.justification = "right",
    legend.margin = margin(6, 0, 1.5, 0, "pt"),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(0, "pt"),
    legend.box.spacing = grid::unit(0, "pt")
  )
```

Turn off clipping.

```{r}
ggplot(data_health, aes(x = age, y = stat(count))) +
  geom_density(
    data = select(data_health, -health),
    aes(fill = "all people surveyed   "),
    color = NA
  ) +
  geom_density(aes(fill = "highlighted group"), color = NA) +
  scale_x_continuous(
    name = "age (years)",
    limits = c(15, 98),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "count",
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#2b8cbed0"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  facet_wrap(~health, nrow = 1) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(12) +
  theme(
    axis.line = element_blank(),
    strip.text = element_text(size = 12, margin = margin(0, 0, 6, 0, "pt")),
    legend.position = "bottom",
    legend.justification = "right",
    legend.margin = margin(6, 0, 1.5, 0, "pt"),
    legend.spacing.x = grid::unit(3, "pt"),
    legend.spacing.y = grid::unit(0, "pt"),
    legend.box.spacing = grid::unit(0, "pt")
  )
```
